---
title: "RGGI Data Analysis"
author: "Eric Scheier"
date: "3/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
```

```{r}
read_then_csv <- function(sheet, path) {
  pathbase <- path %>%
    basename() %>%
    tools::file_path_sans_ext()
  path %>%
    read_excel(sheet = sheet, col_names = FALSE, col_types = "text") %>% 
    write_csv(paste0("data/",sheet, ".csv"))
}

extract_all_sheets <- function(path){
  path %>%
    readxl::excel_sheets() %>%
    set_names() %>% 
    map(read_then_csv, path = path)
}

download_current_egrid <- function(url=NULL){
  temp <- tempfile()
  download.file(url,temp,mode="wb", method="libcurl")
  # mode = 'wb'
  # path <- temp
  extract_all_sheets(path=temp)
}

get_state_egrid <- function(year=2019, relevant_file_path=NULL){
  if(is.null(relevant_file_path)){
    # https://stackoverflow.com/questions/44009924/getting-last-two-digits-from-numeric-in-r
    year_code <- format(as.Date(sprintf('%d-01-01', year)), '%y')
    file_pattern <- paste0("*.ST",year_code,".csv") #".*ST[0-9]*.csv"
    relevant_file_path <- list.files(path="data",pattern=file_pattern,
                                     full.names=TRUE,ignore.case=TRUE)
  }
  message("reading ",relevant_file_path)
  # if(!file.exists(relevant_file_path)){
  #   download_current_egrid()
  # }
  retry <- TRUE
  skip_to_line <- 0
  while(retry){
    state_egrid <- read_csv(relevant_file_path, skip=skip_to_line)
    fips_row <- which(apply(state_egrid, 1, function(x) {any(grepl(x, pattern = "FIPS")) }))
    fips_row <- fips_row[length(fips_row)]
    if(length(fips_row)>0 && fips_row != skip_to_line){
      skip_to_line <- fips_row
    } else {
      retry <- FALSE
    }
  }
  
  names(state_egrid) <- sub("\n.*","",names(state_egrid))
  
  return(state_egrid)
}

download_historical_egrid <- function(url=NULL){
  temp <- tempfile()
  download.file(url,temp)
  unzip(zipfile=temp,exdir="data")
  unlink(temp)
}
```

```{r}
historical_data_url <- "https://www.epa.gov/sites/production/files/2020-01/egrid2018_historical_files_since_1996.zip"
data_url_2018 <- "https://www.epa.gov/sites/production/files/2020-03/egrid2018_data_v2.xlsx"
data_url_2019 <- "https://www.epa.gov/sites/production/files/2021-02/egrid2019_data.xlsx"
```

We are interested in understanding the CO2 emissions intensity of electricity generated and consumed in member states of the Regional Greenhouse Gas Initiative (RGGI) before and during their participation in RGGI.

The data for making this assessment is available from the Environmental Protection Agency (EPA) through its Emissions & Generation Resource Integrated Database (eGRID).

```{r}
if (!dir.exists("data")) {dir.create("data")}

download_current_egrid(url=data_url_2019)
download_current_egrid(url=data_url_2018)
download_historical_egrid(url=historical_data_url)
```

```{r}
historical_files <- list.files(path="data",
                               pattern="^egrid.*?_(location|data|aggregation)(|_v2)\\.xls",
                               full.names=TRUE,ignore.case=TRUE)

lapply(historical_files, extract_all_sheets)
```

```{r}
state_files <- list.files(path="data",
           pattern=".*ST[0-9]*.csv",
           full.names = TRUE,ignore.case=TRUE)
```


We are interested in the following variables for our analysis:

+ PSTATABB: The state abbreviation for easy sorting and viewing
+ YEAR: which year the data was collected by the EPA (we must infer this from the data filenames)
+ STNGENAN: State annual net generation (MWh)
+ STCO2AN: State annual CO2 emissions (tons)
+ STCO2RTA: State average annual CO2 output emission rate (lbs/MWh)

```{r}
desired_columns <- c("state"="PSTATABB",
                     "generation_MWh"="STNGENAN",
                     "emissions_tons"="STCO2AN",
                     "emission_rate_lbs_per_MWh"="STCO2RTA")
```

```{r}
# function to get the year from the filename
year_from_file <- function(file_name){
  current_year <- as.numeric(format(Sys.Date(),"%y"))
  file_year <- as.numeric(gsub("[^\\d]+", "", file_name, perl=TRUE))
  full_file_year <- ifelse(file_year>current_year,1900+file_year,2000+file_year)
  return(full_file_year)
}
```

```{r}
# https://stackoverflow.com/questions/63808658/combining-multiple-csv-files-together-in-an-r-loop
new_data <- purrr::map_df(state_files, function(x) {
  mydata <- get_state_egrid(year=NULL,
                            relevant_file_path = x)
  mydata <- mydata[,desired_columns]
  mydata %>%
    mutate(year = year_from_file(x))
}) %>% 
  rename(all_of(desired_columns))
```

```{r}
write_csv(new_data,"egrid_selected_data_all_years.csv")
```


